\pkg{shorts}: An R Package for Modeling Short Sprints

### 2020-10-21 paper.tex
Mladen changes

\(W\) in equation \eqref{eq:velocity-time} represents Lambert's W function \citep{R-LambertW}. MSS and TAU parameters are estimated using \texttt{shorts::model\_using\_splits} function:

\begin{CodeChunk}
\begin{CodeInput}
R> require(shorts)
R> 
R> split_distance <- c(10, 20, 30, 40)
R> 
R> split_time <- c(2.17, 3.43, 4.60, 5.73)
R> 
R> m1 <- shorts::model_using_splits(
+   distance = split_distance,
+   time = split_time
+ )
R> 
R> m1
\end{CodeInput}
\begin{CodeOutput}
#> Estimated model parameters
#> --------------------------
#>                 MSS                 TAU                 MAC 
#>                9.01                1.31                6.89 
#>                PMAX     time_correction distance_correction 
#>               15.52                0.00                0.00 
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>   0.00249   1.00000  -0.00178   0.00265   0.00265   0.00176   0.00157 
#>      MAPE 
#>   0.04742
\end{CodeOutput}
\end{CodeChunk}

@@

Once we have estimated MSS and TAU, we can use \texttt{shorts::predict\_} family of functions to predict various relationships (i.e.~time at distance, acceleration at distance, velocity at time):

\begin{CodeChunk}
\begin{CodeInput}
R> # Predict time at distance
R> shorts::predict_time_at_distance(
+   distance = split_distance,
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 2.17 3.43 4.60 5.73
\end{CodeOutput}
\begin{CodeInput}
R> 
R> # Predict acceleration at time
R> shorts::predict_acceleration_at_time(
+   time = c(0, 1, 2, 3, 4, 5, 6),
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 6.8884 3.2075 1.4935 0.6954 0.3238 0.1508 0.0702
\end{CodeOutput}
\end{CodeChunk}

Here is a plot of observed and model predicted split times, with the help of the \pkg{ggplot2} package \citep{R-ggplot2}:

\begin{CodeChunk}
\begin{CodeInput}
R> require(ggplot2)
R> 
R> df <- data.frame(
+   distance = split_distance,
+   time = split_time,
+   pred_time = shorts::predict_time_at_distance(
+     distance = split_distance,
+     MSS = m1$parameters$MSS,
+     TAU = m1$parameters$TAU
+   )
+ )
R> 
R> ggplot(df, aes(x = distance)) +
+   theme_bw(8) +
+   geom_line(aes(y = time)) +
+   geom_point(aes(y = pred_time), color = "red") +
+   xlab("Distance (m)") +
+   ylab("Time (s)")
\end{CodeInput}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-4-1} \end{center}

@@

To estimate force production at distance or time (using \texttt{shorts::predict\_force\_at\_distance} and \texttt{shorts::predict\_force\_at\_time} functions), as well as power production (using \texttt{shorts::predict\_power\_at\_distance} and \texttt{shorts::predict\_power\_at\_time} functions), one needs to take into account the air resistance. Air resistance (N) is estimated using \texttt{shorts::get\_air\_resistance} function, which takes velocity, body mass (kg), body height (m), barometric pressure (Torr), air temperature (\(C^\circ\)), and wind velocity (\(ms^-1\)) as parameters (please refer to \citet{arsacModelingEnergetics100m2002}, \citet{samozinoSimpleMethodMeasuring2016}, and \citet{vaningenschenauCanCyclePower1991} for more information):

\begin{CodeChunk}
\begin{CodeInput}
R> shorts::get_air_resistance(
+   velocity = 5,
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 6.1
\end{CodeOutput}
\end{CodeChunk}

When estimating force and power, the air resistance parameters can be set using \texttt{"..."}, which are forwarded to the \texttt{shorts::get\_air\_resistance}:

\begin{CodeChunk}
\begin{CodeInput}
R> # To calculate horizontal force produced
R> shorts::predict_force_at_distance(
+   distance = split_distance,
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU,
+   # Additional parameters forwarded to shorts::get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 119.0  58.6  36.9  28.2
\end{CodeOutput}
\begin{CodeInput}
R> 
R> # To calculate power produced
R> shorts::predict_power_at_distance(
+   distance = split_distance,
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU,
+   # Additional parameters forwarded to shorts::get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 868 490 323 251
\end{CodeOutput}
\end{CodeChunk}

The easiest way to get all kinematics for short sprints is to use \texttt{shorts::predict\_kinematics} functions:

\begin{CodeChunk}
\begin{CodeInput}
R> df <- shorts::predict_kinematics(
+   m1,
+   max_time = 6,
+   frequency = 100,
+   # Additional parameters forwarded to shorts::get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
R> 
R> head(df)
\end{CodeInput}
\begin{CodeOutput}
#>   time distance velocity acceleration air_resistance force power
#> 1 0.00 0.000000   0.0000         6.89        0.07536   551   0.0
#> 2 0.01 0.000344   0.0686         6.84        0.05609   547  37.5
#> 3 0.02 0.001371   0.1367         6.78        0.03978   543  74.2
#> 4 0.03 0.003076   0.2043         6.73        0.02636   539 110.0
#> 5 0.04 0.005455   0.2714         6.68        0.01576   534 145.0
#> 6 0.05 0.008502   0.3379         6.63        0.00792   530 179.2
#>   relative_power
#> 1          0.000
#> 2          0.469
#> 3          0.928
#> 4          1.375
#> 5          1.813
#> 6          2.241
\end{CodeOutput}
\end{CodeChunk}

Plotting the model predictions can be done once we convert data from wide to long with the help of \pkg{dplyr} \citep{R-dplyr}, \pkg{tidyr} \citep{R-tidyr}, and \pkg{tidyverse} \citep{R-tidyverse} packages:

@@

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-8-1} \end{center}

@@

Another valuable addition for sport scientists and coaches is the ability to determine the distances and times where 90\% of maximum sprinting speed is reached, or where peak power is within 90\% range. To identify these values, \pkg{shorts} package comes with \texttt{shorts::find\_} family of functions:

\begin{CodeChunk}
\begin{CodeInput}
R> # Finds distance where 90% of maximum sprinting speed is reached
R> shorts::find_velocity_critical_distance(
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU,
+   percent = 0.9
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 16.5
\end{CodeOutput}
\begin{CodeInput}
R> 
R> # Finds maximal power and distance (this time using air resistance)
R> shorts::find_max_power_distance(
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU,
+   # Additional parameters forwarded to shorts::get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
\end{CodeInput}
\begin{CodeOutput}
#> $max_power
#> [1] 1264
#> 
#> $distance
#> [1] 2.46
\end{CodeOutput}
\begin{CodeInput}
R> 
R> 
R> # Finds distance over 90% power range
R> shorts::find_power_critical_distance(
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU,
+   # Additional parameters forwarded to shorts::get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
\end{CodeInput}
\begin{CodeOutput}
#> $lower
#> [1] 0.959
#> 
#> $upper
#> [1] 5.44
\end{CodeOutput}
\end{CodeChunk}

@@

Sprint performance is often evaluated with a group of athletes (e.g., soccer club) representing a single strata of interest. Sports scientists can estimate individual profiles, or utilize mixed-effects models. To perform mixed-effects models in \pkg{shorts} for split times, one can use \texttt{shorts::mixed\_model\_using\_splits} function. To demonstrate this functionality, we load the \texttt{split\_times} dataset provided in the \pkg{shorts} package:

\begin{CodeChunk}
\begin{CodeInput}
R> data(split_times)
R> 
R> head(split_times)
\end{CodeInput}
\begin{CodeOutput}
#> # A tibble: 6 x 4
#>   athlete bodyweight distance     time
#>   <chr>        <dbl>    <dbl> <I<dbl>>
#> 1 John            75        5     1.20
#> 2 John            75       10     1.97
#> 3 John            75       15     2.66
#> 4 John            75       20     3.31
#> 5 John            75       30     4.59
#> 6 John            75       40     5.85
\end{CodeOutput}
\begin{CodeInput}
R> 
R> # Mixed model
R> m2 <- shorts::mixed_model_using_splits(
+   data = split_times,
+   distance = "distance",
+   time = "time",
+   athlete = "athlete",
+ 
+   # Select random effects
+   # Default is MSS and TAU
+   random = MSS + TAU ~ 1
+ )
R> 
R> m2
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>               8.065               0.655              12.309 
#>                PMAX     time_correction distance_correction 
#>              24.818               0.000               0.000 
#> 
#> Estimated random model parameters
#> ----------------------------------
#>     athlete  MSS   TAU  MAC PMAX time_correction distance_correction
#> 1     James 9.69 0.847 11.4 27.7               0                   0
#> 2       Jim 7.83 0.505 15.5 30.4               0                   0
#> 3      John 7.78 0.727 10.7 20.8               0                   0
#> 4 Kimberley 8.57 0.802 10.7 22.9               0                   0
#> 5  Samantha 6.45 0.395 16.3 26.4               0                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>    0.0260    0.9998   -0.0293    0.0496    0.0496    0.0214    0.0172 
#>      MAPE 
#>    0.9019
\end{CodeOutput}
\end{CodeChunk}

@@

The following plot contains kinematics for all athletes in \texttt{split\_times} dataset. Please note that power calculation takes default parameters for each individual:

\begin{CodeChunk}
\begin{CodeInput}
R> df <- predict_kinematics(m2, max_time = 10)
R> 
R> df <- pivot_longer(df, cols = c(-1, -3))
R> 
R> ggplot(
+   filter(df, distance < 40),
+   aes(x = distance, y = value, group = athlete, color = athlete)
+ ) +
+   theme_bw(8) +
+   facet_wrap(~name, scales = "free_y") +
+   geom_line(alpha = 0.7) +
+   ylab(NULL) +
+   xlab("Distance (m)") +
+   theme(legend.position = "top", legend.title = element_blank())
\end{CodeInput}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-12-1} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> sprint_time <- seq(0, 6, 1)
R> 
R> sprint_velocity <- c(0.00, 4.83, 7.07, 8.10, 8.59, 8.81, 8.91)
R> 
R> m3 <- shorts::model_using_radar(
+   velocity = sprint_velocity,
+   time = sprint_time
+ )
R> 
R> m3
\end{CodeInput}
\begin{CodeOutput}
#> Estimated model parameters
#> --------------------------
#>                 MSS                 TAU                 MAC 
#>                9.00                1.30                6.92 
#>                PMAX     time_correction distance_correction 
#>               15.58                0.00                0.00 
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>   0.00327   1.00000  -0.00406   0.00532   0.00532   0.00276   0.00207 
#>      MAPE 
#>       NaN
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> m3_weighted <- shorts::model_using_radar(
+   velocity = sprint_velocity,
+   time = sprint_time,
+   weights = 1 / (sprint_velocity + 1)
+ )
R> 
R> m3_weighted
\end{CodeInput}
\begin{CodeOutput}
#> Estimated model parameters
#> --------------------------
#>                 MSS                 TAU                 MAC 
#>                9.00                1.30                6.92 
#>                PMAX     time_correction distance_correction 
#>               15.58                0.00                0.00 
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>   0.00108   1.00000  -0.00406   0.00534   0.00534   0.00276   0.00206 
#>      MAPE 
#>       NaN
\end{CodeOutput}
\end{CodeChunk}

@@

Mixed-effects model using radar data is done using \texttt{shorts::mixed\_model\_using\_radar} function. To perform mixed model, let's load data that comes with \pkg{shorts} package.

\begin{CodeChunk}
\begin{CodeInput}
R> data("radar_gun_data")
R> 
R> head(radar_gun_data)
\end{CodeInput}
\begin{CodeOutput}
#> # A tibble: 6 x 4
#>   athlete bodyweight  time velocity
#>   <chr>        <dbl> <dbl>    <dbl>
#> 1 John            75  0       0    
#> 2 John            75  0.01    0.075
#> 3 John            75  0.02    0.149
#> 4 John            75  0.03    0.222
#> 5 John            75  0.04    0.291
#> 6 John            75  0.05    0.367
\end{CodeOutput}
\begin{CodeInput}
R> 
R> m4 <- shorts::mixed_model_using_radar(
+   radar_gun_data,
+   time = "time",
+   velocity = "velocity",
+   athlete = "athlete"
+ )
R> 
R> m4
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>                8.30                1.01                8.24 
#>                PMAX     time_correction distance_correction 
#>               17.09                0.00                0.00 
#> 
#> Estimated random model parameters
#> ----------------------------------
#>     athlete   MSS   TAU  MAC PMAX time_correction distance_correction
#> 1     James 10.00 1.111 9.00 22.5               0                   0
#> 2       Jim  8.00 0.889 9.00 18.0               0                   0
#> 3      John  8.00 1.069 7.48 15.0               0                   0
#> 4 Kimberley  9.01 1.286 7.01 15.8               0                   0
#> 5  Samantha  6.50 0.685 9.50 15.4               0                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>    0.0516    0.9994   -0.2191    0.1983    0.2191    0.0516    0.0395 
#>      MAPE 
#>       NaN
\end{CodeOutput}
\end{CodeChunk}

@@

One source of error in the modeled estimation using a radar gun is the time synchronization. In theory, synchronization is ideal when a sprint is initiated at \(t=0\) (i.e., \(v(t=0) = 0\)). In practice, this is often not the case. Let's use our athlete and add and deduct 0.5s to simulate an error in synchronization and its effect on estimated MSS and TAU.

\begin{CodeChunk}
\begin{CodeInput}
R> df <- tibble(
+   `true time` = sprint_time,
+   velocity = sprint_velocity,
+   `0.5s added` = `true time` + 0.5,
+   `0.5s deducted` = `true time` - 0.5
+ )
R> 
R> head(df)
\end{CodeInput}
\begin{CodeOutput}
#> # A tibble: 6 x 4
#>   `true time` velocity `0.5s added` `0.5s deducted`
#>         <dbl>    <dbl>        <dbl>           <dbl>
#> 1           0     0             0.5            -0.5
#> 2           1     4.83          1.5             0.5
#> 3           2     7.07          2.5             1.5
#> 4           3     8.1           3.5             2.5
#> 5           4     8.59          4.5             3.5
#> 6           5     8.81          5.5             4.5
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> plot_df <- pivot_longer(df, cols = -2, names_to = "Sync issue")
R> 
R> ggplot(plot_df, aes(x = value, y = velocity, color = `Sync issue`)) +
+   theme_bw(8) +
+   geom_line(alpha = 0.7) +
+   xlab("Time (s)") +
+   ylab(expression("Velocity (" * ms^-1 * ")"))
\end{CodeInput}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-17-1} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # Without synchronization issues
R> m5 <- shorts::model_using_radar(
+   velocity = df$velocity,
+   time = df$`true time`
+ )
R> 
R> # With time added
R> m6 <- shorts::model_using_radar(
+   velocity = df$velocity,
+   time = df$`0.5s added`
+ )
R> 
R> # With time deducted
R> m7 <- shorts::model_using_radar(
+   velocity = df$velocity,
+   time = df$`0.5s deducted`
+ )
R> 
R> rbind(
+   data.frame(
+     model = "True time",
+     t(coef(m5))
+   ),
+   data.frame(
+     model = "Added 0.5s time",
+     t(coef(m6))
+   ),
+   data.frame(
+     model = "Deducted 0.5s time",
+     t(coef(m7))
+   )
+ )
\end{CodeInput}
\begin{CodeOutput}
#>                model   MSS  TAU  MAC PMAX time_correction
#> 1          True time  9.00 1.30 6.92 15.6               0
#> 2    Added 0.5s time  9.91 2.34 4.23 10.5               0
#> 3 Deducted 0.5s time 10.08 1.86 5.43 13.7               0
#>   distance_correction
#> 1                   0
#> 2                   0
#> 3                   0
\end{CodeOutput}
\end{CodeChunk}

@@

This model is incorporated in the \texttt{shorts::model\_using\_radar\_with\_time\_correction} function:

>>>>

\(W\) in equation \eqref{eq:velocity-time} represents Lambert's W function \citep{R-LambertW}. MSS and TAU parameters are estimated using \texttt{model\_using\_splits} function:

\begin{CodeChunk}
\begin{CodeInput}
R> require(shorts)
R> 
R> split_distance <- c(10, 20, 30, 40)
R> 
R> split_time <- c(2.17, 3.43, 4.60, 5.73)
R> 
R> m1 <- model_using_splits(
+   distance = split_distance,
+   time = split_time
+ )
R> 
R> m1
\end{CodeInput}
\begin{CodeOutput}
#> Estimated model parameters
#> --------------------------
#>                 MSS                 TAU                 MAC 
#>                9.01                1.31                6.89 
#>                PMAX     time_correction distance_correction 
#>               15.52                0.00                0.00 
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>   0.00249   1.00000  -0.00178   0.00265   0.00265   0.00176   0.00157 
#>      MAPE 
#>   0.04742
\end{CodeOutput}
\end{CodeChunk}

@@

Once we have estimated MSS and TAU, we can use \texttt{predict\_} family of functions to predict various relationships (i.e.~time at distance, acceleration at distance, velocity at time):

\begin{CodeChunk}
\begin{CodeInput}
R> # Predict time at distance
R> predict_time_at_distance(
+   distance = split_distance,
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 2.17 3.43 4.60 5.73
\end{CodeOutput}
\begin{CodeInput}
R> 
R> # Predict acceleration at time
R> predict_acceleration_at_time(
+   time = c(0, 1, 2, 3, 4, 5, 6),
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 6.8884 3.2075 1.4935 0.6954 0.3238 0.1508 0.0702
\end{CodeOutput}
\end{CodeChunk}

@@

To estimate force production at distance or time (using \texttt{predict\_force\_at\_distance} and \texttt{predict\_force\_at\_time} functions), as well as power production (using \texttt{predict\_power\_at\_distance} and \texttt{predict\_power\_at\_time} functions), one needs to take into account the air resistance. Air resistance (N) is estimated using \texttt{get\_air\_resistance} function, which takes velocity, body mass (kg), body height (m), barometric pressure (Torr), air temperature (\(C^\circ\)), and wind velocity (\(ms^-1\)) as parameters (please refer to \citet{arsacModelingEnergetics100m2002}, \citet{samozinoSimpleMethodMeasuring2016}, and \citet{vaningenschenauCanCyclePower1991} for more information):

\begin{CodeChunk}
\begin{CodeInput}
R> get_air_resistance(
+   velocity = 5,
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 6.1
\end{CodeOutput}
\end{CodeChunk}

When estimating force and power, the air resistance parameters can be set using \texttt{"..."}, which are forwarded to the \texttt{get\_air\_resistance}:

\begin{CodeChunk}
\begin{CodeInput}
R> # To calculate horizontal force produced
R> predict_force_at_distance(
+   distance = split_distance,
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU,
+   # Additional parameters forwarded to get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 119.0  58.6  36.9  28.2
\end{CodeOutput}
\begin{CodeInput}
R> 
R> # To calculate power produced
R> predict_power_at_distance(
+   distance = split_distance,
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU,
+   # Additional parameters forwarded to get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 868 490 323 251
\end{CodeOutput}
\end{CodeChunk}

The easiest way to get all kinematics for short sprints is to use \texttt{predict\_kinematics} functions:

\begin{CodeChunk}
\begin{CodeInput}
R> df <- predict_kinematics(
+   m1,
+   max_time = 6,
+   frequency = 100,
+   # Additional parameters forwarded to get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
R> 
R> head(df)
\end{CodeInput}
\begin{CodeOutput}
#>   time distance velocity acceleration air_resistance force power
#> 1 0.00 0.000000   0.0000         6.89        0.07536   551   0.0
#> 2 0.01 0.000344   0.0686         6.84        0.05609   547  37.5
#> 3 0.02 0.001371   0.1367         6.78        0.03978   543  74.2
#> 4 0.03 0.003076   0.2043         6.73        0.02636   539 110.0
#> 5 0.04 0.005455   0.2714         6.68        0.01576   534 145.0
#> 6 0.05 0.008502   0.3379         6.63        0.00792   530 179.2
#>   relative_power
#> 1          0.000
#> 2          0.469
#> 3          0.928
#> 4          1.375
#> 5          1.813
#> 6          2.241
\end{CodeOutput}
\end{CodeChunk}

Plotting the model predictions can be done once we convert data from wide to long with the help of \pkg{ggplot2} \citep{R-ggplot2}, \pkg{dplyr} \citep{R-dplyr}, \pkg{tidyr} \citep{R-tidyr}, and \pkg{tidyverse} \citep{R-tidyverse} packages:

@@

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-7-1} \end{center}

@@

Another valuable addition for sport scientists and coaches is the ability to determine the distances and times where 90\% of maximum sprinting speed is reached, or where peak power is within 90\% range. To identify these values, \pkg{shorts} package comes with \texttt{find\_} family of functions:

\begin{CodeChunk}
\begin{CodeInput}
R> # Finds distance where 90% of maximum sprinting speed is reached
R> find_velocity_critical_distance(
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU,
+   percent = 0.9
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 16.5
\end{CodeOutput}
\begin{CodeInput}
R> 
R> # Finds maximal power and distance (this time using air resistance)
R> find_max_power_distance(
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU,
+   # Additional parameters forwarded to get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
\end{CodeInput}
\begin{CodeOutput}
#> $max_power
#> [1] 1264
#> 
#> $distance
#> [1] 2.46
\end{CodeOutput}
\begin{CodeInput}
R> 
R> # Finds distance over 90% power range
R> find_power_critical_distance(
+   MSS = m1$parameters$MSS,
+   TAU = m1$parameters$TAU,
+   # Additional parameters forwarded to get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
\end{CodeInput}
\begin{CodeOutput}
#> $lower
#> [1] 0.959
#> 
#> $upper
#> [1] 5.44
\end{CodeOutput}
\end{CodeChunk}

@@

Sprint performance is often evaluated with a group of athletes (e.g., soccer club) representing a single strata of interest. Sports scientists can estimate individual profiles, or utilize mixed-effects models. To perform mixed-effects models in \pkg{shorts} for split times, one can use \texttt{mixed\_model\_using\_splits} function. To demonstrate this functionality, we load the \texttt{split\_times} dataset provided in the \pkg{shorts} package:

\begin{CodeChunk}
\begin{CodeInput}
R> data(split_times)
R> 
R> # Mixed model
R> m2 <- mixed_model_using_splits(
+   data = split_times,
+   distance = "distance",
+   time = "time",
+   athlete = "athlete",
+ 
+   # Select random effects
+   # Default is MSS and TAU
+   random = MSS + TAU ~ 1
+ )
R> 
R> m2
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>               8.065               0.655              12.309 
#>                PMAX     time_correction distance_correction 
#>              24.818               0.000               0.000 
#> 
#> Estimated random model parameters
#> ----------------------------------
#>     athlete  MSS   TAU  MAC PMAX time_correction distance_correction
#> 1     James 9.69 0.847 11.4 27.7               0                   0
#> 2       Jim 7.83 0.505 15.5 30.4               0                   0
#> 3      John 7.78 0.727 10.7 20.8               0                   0
#> 4 Kimberley 8.57 0.802 10.7 22.9               0                   0
#> 5  Samantha 6.45 0.395 16.3 26.4               0                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>    0.0260    0.9998   -0.0293    0.0496    0.0496    0.0214    0.0172 
#>      MAPE 
#>    0.9019
\end{CodeOutput}
\end{CodeChunk}

@@

The following figure contains kinematics for all athletes in \texttt{split\_times} dataset. Please note that power calculation takes default parameters for each individual:

\begin{CodeChunk}
\begin{CodeInput}
R> df <- predict_kinematics(m2, max_time = 10)
R> 
R> df <- pivot_longer(df, cols = c(-1, -3))
R> 
R> ggplot(
+   filter(df, distance < 40),
+   aes(x = distance, y = value, group = athlete, color = athlete)
+ ) +
+   theme_bw(8) +
+   facet_wrap(~name, scales = "free_y") +
+   geom_line(alpha = 0.7) +
+   ylab(NULL) +
+   xlab("Distance (m)") +
+   theme(
+     legend.position = "top",
+     legend.title = element_blank())
\end{CodeInput}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-11-1} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> sprint_time <- seq(0, 6, 1)
R> 
R> sprint_velocity <- c(0.00, 4.83, 7.07, 8.10, 8.59, 8.81, 8.91)
R> 
R> m3 <- model_using_radar(
+   velocity = sprint_velocity,
+   time = sprint_time
+ )
R> 
R> m3
\end{CodeInput}
\begin{CodeOutput}
#> Estimated model parameters
#> --------------------------
#>                 MSS                 TAU                 MAC 
#>                9.00                1.30                6.92 
#>                PMAX     time_correction distance_correction 
#>               15.58                0.00                0.00 
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>   0.00327   1.00000  -0.00406   0.00532   0.00532   0.00276   0.00207 
#>      MAPE 
#>       NaN
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> m3_weighted <- model_using_radar(
+   velocity = sprint_velocity,
+   time = sprint_time,
+   weights = 1 / (sprint_velocity + 1)
+ )
R> 
R> m3_weighted
\end{CodeInput}
\begin{CodeOutput}
#> Estimated model parameters
#> --------------------------
#>                 MSS                 TAU                 MAC 
#>                9.00                1.30                6.92 
#>                PMAX     time_correction distance_correction 
#>               15.58                0.00                0.00 
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>   0.00108   1.00000  -0.00406   0.00534   0.00534   0.00276   0.00206 
#>      MAPE 
#>       NaN
\end{CodeOutput}
\end{CodeChunk}

@@

Mixed-effects model using radar data is done using \texttt{mixed\_model\_using\_radar} function. To perform mixed model, let's load data that comes with \pkg{shorts} package.

\begin{CodeChunk}
\begin{CodeInput}
R> data("radar_gun_data")
R> 
R> m4 <- mixed_model_using_radar(
+   radar_gun_data,
+   time = "time",
+   velocity = "velocity",
+   athlete = "athlete"
+ )
R> 
R> m4
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>                8.30                1.01                8.24 
#>                PMAX     time_correction distance_correction 
#>               17.09                0.00                0.00 
#> 
#> Estimated random model parameters
#> ----------------------------------
#>     athlete   MSS   TAU  MAC PMAX time_correction distance_correction
#> 1     James 10.00 1.111 9.00 22.5               0                   0
#> 2       Jim  8.00 0.889 9.00 18.0               0                   0
#> 3      John  8.00 1.069 7.48 15.0               0                   0
#> 4 Kimberley  9.01 1.286 7.01 15.8               0                   0
#> 5  Samantha  6.50 0.685 9.50 15.4               0                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>    0.0516    0.9994   -0.2191    0.1983    0.2191    0.0516    0.0395 
#>      MAPE 
#>       NaN
\end{CodeOutput}
\end{CodeChunk}

@@

One source of error in the modeled estimation using a radar gun is the time synchronization. In theory, synchronization is ideal when a sprint is initiated at \(t=0\) (i.e., \(v(t=0) = 0\)). In practice, this is often not the case. Let's use our athlete and add and deduct 0.5s to simulate an error in synchronization and its effect on estimated MSS and TAU (Figure \ref{fig:radar-gun-sync}).

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # Without synchronization issues
R> m5 <- model_using_radar(
+   velocity = df$velocity,
+   time = df$`true time`
+ )
R> 
R> # With time added
R> m6 <- model_using_radar(
+   velocity = df$velocity,
+   time = df$`0.5s added`
+ )
R> 
R> # With time deducted
R> m7 <- model_using_radar(
+   velocity = df$velocity,
+   time = df$`0.5s deducted`
+ )
R> 
R> rbind(
+   data.frame(
+     model = "True time",
+     t(coef(m5))
+   ),
+   data.frame(
+     model = "Added 0.5s time",
+     t(coef(m6))
+   ),
+   data.frame(
+     model = "Deducted 0.5s time",
+     t(coef(m7))
+   )
+ )
\end{CodeInput}
\begin{CodeOutput}
#>                model   MSS  TAU  MAC PMAX time_correction
#> 1          True time  9.00 1.30 6.92 15.6               0
#> 2    Added 0.5s time  9.91 2.34 4.23 10.5               0
#> 3 Deducted 0.5s time 10.08 1.86 5.43 13.7               0
#>   distance_correction
#> 1                   0
#> 2                   0
#> 3                   0
\end{CodeOutput}
\end{CodeChunk}

@@

This model is incorporated in the \texttt{model\_using\_radar\_with\_time\_correction} function:
